Type-Safe Modular Parsing

### 2017-01-14 paper/sections/Introduction.tex
More space tweak
S1.

To address the modularity limitations of Programming Languages, several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice, due to their simplicity of implementation and use.
Examples include many tools for developing Feature-Oriented Software-Product
Lines (SPLs)~\cite{Kastner11road}, some Language Workbenches~\cite{}, or extensible parser
generators~\cite{}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{} to
enable the development modular program features. Such
composition techniques collect the code for multiple features and
merge them together when a concrete combination of features is needed
for a particular program. As Kastner et. al~\cite{Kastner11road} note,
a typical drawback of feature-oriented SPL implementations, which
more generally applies to syntactic modularity approaches, is that
such ``\emph{implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''.

@@

  To address such algorithmic challanges, we propose a methodology
  for implementing modular parsers built on top of an
  existing Packrat~\cite{Ford2002} parsing library for Scala~\cite{}. The Packrat parsing library
  directly supports left-recursion, and a \emph{longest-match
    composition} operator. The direct support for left-recursion
  avoids the need for left-recursion elimination, which is very
  cumbersome is a modular setting. The longest-match composition
  operator avoids the use of priorities, and it provides reasonably
  efficient optimizations for backtracking.

  \paragraph{Typing and Reusability Challenges} The second class of
  challenges was problems related to modularity, reusability and
  typing of parsing code. An immediate concern is how to extend a
  parser for an existing language or, more generally, how to compose
  parsing code for two languages. It turns out that OO mechanisms that
  provide some form of multiple inheritance,
  such as traits/mixins~\cite{scalaoverview}\bruno{cite work on traits (Ducasse)
  and mixins (Bracha)}, are very handy for this
  purpose. Essentially, traits/mixins can act as modules for the parsing code
  of different languages. This enables an approach where ASTs can be
  modelled using standard OO techniques such as the {\sc Composite}
  pattern, while retaining the possibility of adding new language
  constructs. Section~\ref{sec:inheritance} gives the details of this approach.

@@

To evaluate our approach we conduct a case study based on the ``Types
and Programming Languages'' (TAPL) interpreters. The case study shows
that our approach is effective at reusing parsing code from existing
interpreters, and the total parsing code is 69\% shorter than an
existing code base using non-modular parsing code~\cite{}.

@@

\begin{itemize}

@@

All the code in the paper and the case study is written in Scala. But,
other languages that support some form of multiple inheritance
(including Java 8 with default methods~\cite{}) could in principle be
used. Our choice for Scala had to do with the concise and elegant syntax
of the language which enables a clear presentation of our ideas.

>>>>

To address the modularity limitations of Programming Languages, several
different approaches have been proposed in the past. Existing
approaches can be broadly divided into two categories:
\emph{syntactic} or \emph{semantic} modularization
techniques. Syntactic modularization techniques are quite popular in
practice, due to their simplicity of implementation and use.
Examples include many tools for developing Feature-Oriented Software-Product
Lines (SPLs)~\cite{Kastner11road}, some Language Workbenches~\cite{Erdweg201524}, or extensible parser
generators~\cite{antlr1995,Grimm2006,Gouseti2014,Warth2016}.  Most syntactic approaches employ textual
composition techniques such as \emph{superimposition}~\cite{Wagelaar2010} to
enable the development modular program features. Such
composition techniques collect the code for multiple features and
merge them together when a concrete combination of features is needed
for a particular program. As Kastner et. al~\cite{Kastner11road} note,
a typical drawback of feature-oriented SPL implementations, which
more generally applies to syntactic modularity approaches, is that
such ``\emph{implementation mechanisms lack proper
  interfaces and support neither modular type checking nor separate
  compilation}''.

@@

  To address such algorithmic challanges, we propose a methodology
  for implementing modular parsers built on top of an
  existing Packrat~\cite{Ford2002} parsing library for Scala~\cite{odersky2004overview}. The Packrat parsing library
  directly supports left-recursion, and a \emph{longest-match
    composition} operator. The direct support for left-recursion
  avoids the need for left-recursion elimination, which is very
  cumbersome is a modular setting. The longest-match composition
  operator avoids the use of priorities, and it provides reasonably
  efficient optimizations for backtracking.

  \paragraph{Typing and Reusability Challenges} The second class of
  challenges was problems related to modularity, reusability and
  typing of parsing code. An immediate concern is how to extend a
  parser for an existing language or, more generally, how to compose
  parsing code for two languages. It turns out that OO mechanisms that
  provide some form of multiple inheritance,
  such as traits/mixins~\cite{Bracha1990,Scharli2003}, are very handy for this
  purpose. Essentially, traits/mixins can act as modules for the parsing code
  of different languages. This enables an approach where ASTs can be
  modelled using standard OO techniques such as the {\sc Composite}
  pattern, while retaining the possibility of adding new language
  constructs. Section~\ref{sec:inheritance} gives the details of this approach.

@@

To evaluate our approach we conduct a case study based on the ``Types
and Programming Languages'' (TAPL) interpreters. The case study shows
that our approach is effective at reusing parsing code from existing
interpreters, and the total parsing code is 69\% shorter than an
existing code base using non-modular parsing code\footnote{https://github.com/ilya-klyuchnikov/tapl-scala/}.

@@

\begin{itemize}[noitemsep,nolistsep]

@@

All the code in the paper and the case study is written in Scala. But,
other languages that support some form of multiple inheritance
(including Java 8 with default methods~\cite{Goetz2012}) could in principle be
used. Our choice for Scala had to do with the concise and elegant syntax
of the language which enables a clear presentation of our ideas.
